Why there are more species in several small patches vs few large patches:

### 2024-03-11 ms_SLOSS.Rmd
Updated discussion

The SLOSS debate has been approached from several theoretical frameworks, but most of them were based on single-species models [@Fahrig2021]. However, recent studies using spatially explicit simulations revealed that while high fragmentation leads to more local extinctions due to the reduced habitat size, at low levels of fragmentation, competitive exclusion is more prevalent, resulting in decreased diversity [@Rybicki2020].

@@

The novel aspect of our model is that it considers communities that are assembled from a metacommunity in which individuals migrate to an empty habitat. After the number of species is stabilized the habitat is fragmented. This scenario is deemed more realistic than having all species initially present with random patterns in an already fragmented habitat  [@Rybicki2020]. This could be more important the more heterogeneous was species spatial distribution pattern before fragmentation, as the cube hypothesis states.

@@

Our findings support Fahrig's [-@Fahrig2021] conclusion that conditions favoring SL > SS are limited and infrequently observed in nature. Under conditions of adequate patch size and dispersal capabilities, the SS scenario typically retains more species. This underscores the ecological significance of smaller habitat patches in maintaining biodiversity and landscape connectivity [@Yan2021; @Manning2006; @Shafer1995]. Our results thus challenge the conventional preference for larger reserves in conservation planning and support the paradigm shift in conservation towards considering sets of many small patches generally having higher biodiversity than sets of few large patches [@Riva2023].

In conclusion, we believe our study adds valuable perspectives to the SLOSS debate, illustrating the complexities of habitat fragmentation and species conservation. While our model has inherent limitations, it provides a useful framework for understanding key ecological processes. Future research should aim to extend these findings to more diverse ecological settings and incorporate additional factors influencing biodiversity in fragmented landscapes.

>>>>

The SLOSS debate has been approached from several theoretical frameworks, but most of them were based on single-species models [@Fahrig2021]. Despite the long-standing discussion, there remains a notable scarcity of multi-species models explicitly designed to validate fragmentation theories. However, recent studies using spatially explicit simulations revealed that while high fragmentation leads to more local extinctions due to the reduced habitat size, at low levels of fragmentation, competitive exclusion is more prevalent, resulting in decreased diversity [@Rybicki2020].

@@

One novel aspect of our model is that it considers communities that are assembled from a metacommunity in which individuals migrate to an empty habitat. After the number of species is stabilized the habitat is fragmented. This scenario is deemed more realistic than having all species initially present with random patterns in an already fragmented habitat  [@Rybicki2020]. This could be more important the more heterogeneous was species spatial distribution pattern before fragmentation, as the cube hypothesis states.

@@

Our findings support Fahrig's [-@Fahrig2021] conclusion that conditions favoring SL > SS are limited and infrequently observed in nature. Under conditions of adequate patch size and dispersal capabilities, the SS scenario typically retains more species. This underscores the ecological significance of smaller habitat patches in maintaining biodiversity and landscape connectivity [@Yan2021; @Manning2006; @Shafer1995]. Our results thus challenge the conventional preference for larger reserves in conservation planning and support the paradigm shift in conservation towards considering sets of many small patches generally having higher biodiversity than sets of few large patches [@Riva2023]. The model's primary outcomes could be validated using empirical data by assessing the prediction that species with greater dispersal capabilities would exhibit a lower SS effect compared to species with limited dispersal.

In conclusion, we believe our study adds valuable perspectives to the SLOSS debate, illustrating the complexities of habitat fragmentation and species conservation. While our model has inherent limitations, it provides a useful framework for understanding key ecological processes. Future research should aim to extend these findings to more diverse ecological settings and incorporate how heterogeneity between species influence biodiversity in fragmented landscapes.
